---
title: "Mapping All Fifty U.S. States"
author: "William Murphy"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mapping All Fifty U.S. States}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This data package provides an easy way to plot 50-state choropleth thematic maps
with ggplot2 in R. 

## Motivation

It is far too difficult to create a U.S. state data visualization in R that
includes Alaska and Hawaii. The U.S. state map included in the `maps` package
includes only the 48 contiguous states. A search for existing
solutions will lead to Stack Overflow threads 
([1](http://stackoverflow.com/questions/13757771/),
[2](http://stackoverflow.com/questions/25530358/)), but the solutions available
require either limiting plotting options by rendering with a specific package,
installing
a large list of dependencies and tracking down shape files to perform 
arcane transformation calculations, 
or using
annotations or visibility-based maps that compromise the aesthetic of the
visualization.

This package is a simple solution: a data file, with no dependencies, that can
be dropped into ggplot2 with no effort. To create it, 
I followed the steps in a 
[Stack Overflow answer](http://stackoverflow.com/a/13767984) 
from Barry Rowlingson to 
transform Alaska and Hawaii into insets next to the contiguous states and
packaged the resulting data frame for simple drop-in use. 

## Usage Examples

To use the fifty state map, simply point the `map` argument in `geom_map`
to `fifty_states`. As usual, you'll also create an aesthetic mapping for 
`map_id` to the column in your data frame that contains the lower
case state names.

If you need to be highly technically accurate, you can add bounding boxes to 
highlight the insets so that no viewer confuses Alaska and Hawaii for regions in
a deformed Mexico that you randomly decided to include. 
I've packaged those up in a `layer`
object called `fifty_states_inset_boxes` (created via `annotate`), so you can 
just add it to your plot. 

```{r example, fig.show='hold'}
library(ggplot2)
library(fiftystater)

data("fifty_states") # this line is optional due to lazy data loading

crimes <- data.frame(state = tolower(rownames(USArrests)), USArrests)

# map_id creates the aesthetic mapping to the state name column in your data
p <- ggplot(crimes, aes(map_id = state)) + 
  # map points to the fifty_states shape data
  geom_map(aes(fill = Assault), map = fifty_states) + 
  expand_limits(x = fifty_states$long, y = fifty_states$lat) +
  coord_map() +
  scale_x_continuous(breaks = NULL) + 
  scale_y_continuous(breaks = NULL) +
  labs(x = "", y = "") +
  theme(legend.position = "bottom", 
        panel.background = element_blank())

p
# add border boxes to AK/HI
p + fifty_states_inset_boxes 
```

This package also plays nicely with my
[colorplaner](https://cran.r-project.org/package=colorplaner)
package for bivariate color scales.

```{r example_colorplaner, fig.width=6}
# Map a second variable to each state's fill color with colorplaner
library(colorplaner)
p + aes(fill2 = UrbanPop) + scale_fill_colorplane() +
  theme(legend.position = "right")
```

The map data included in this package can be generated with the code below. 
If you want to run this code, you
will need to ensure the packages listed below are installed; they are not
automatically
installed with `fiftystater`.

```{r makemap, message=FALSE, eval=FALSE}
# Create map data with AK, HI inset.
library(maptools)
library(rgeos)
library(rgdal)
library(dplyr)

transform_state <- function(object, rot, scale, shift){
  object %>% elide(rotate = rot) %>%
    elide(scale = max(apply(bbox(object), 1, diff)) / scale) %>%
    elide(shift = shift)
}

#state shape file from
# http://www.arcgis.com/home/item.html?id=f7f805eb65eb4ab787a0a3e1116ca7e5
loc <- file.path(tempdir(), "stats_dat")
unzip(system.file("extdata", "states_21basic.zip", package = "fiftystater"),
      exdir = loc)
fifty_states_sp <- readOGR(dsn = loc, layer = "states", verbose = FALSE) %>%
  spTransform(CRS("+init=epsg:2163"))

alaska <- fifty_states_sp[fifty_states_sp$STATE_NAME == "Alaska", ] %>%
  transform_state(-35, 2.5, c(-2400000, -2100000))
proj4string(alaska) <- proj4string(fifty_states_sp)

hawaii <- fifty_states_sp[fifty_states_sp$STATE_NAME == "Hawaii", ] %>%
  transform_state(-35, .75, c(-1170000,-2363000))
proj4string(hawaii) <- proj4string(fifty_states_sp)

fifty_states <-
  fifty_states_sp[!fifty_states_sp$STATE_NAME %in% c("Alaska","Hawaii"), ] %>%
  rbind(alaska) %>%
  rbind(hawaii) %>%
  spTransform(CRS("+init=epsg:4326")) %>%
  fortify(region = "STATE_NAME") %>%
  mutate(id = tolower(id))
```
